---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# getACS

<!-- badges: start -->

[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![License: MIT](https://img.shields.io/badge/License-MIT-yellow.svg)](https://opensource.org/licenses/MIT)
[![Project Status: WIP -- Initial development is in progress, but there has not yet been a stable, usable release suitable for the public.](https://www.repostatus.org/badges/latest/wip.svg)](https://www.repostatus.org/#wip)
[![Codecov test coverage](https://codecov.io/gh/elipousson/getACS/graph/badge.svg)](https://app.codecov.io/gh/elipousson/getACS)
<!-- badges: end -->

The goal of getACS is to make it easier to work with [American Community Survey data](https://www.census.gov/programs-surveys/acs) from the [tidycensus package](https://walker-data.com/tidycensus/) by Kyle Walker and others.

This package includes:

-   Functions that extend `tidycensus::get_acs()` to support multiple tables, geographies, or years
-   Functions for creating formatted tables from ACS data using the [gt package](https://gt.rstudio.com/)

As of April 2024, this package uses a development version of `{tigris}`, available at <https://github.com/elipousson/tigris>.

## Installation

You can install the development version of getACS from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pkg_install("elipousson/getACS")
```

## Usage

```{r setup}
library(getACS)
library(gt)
library(ggplot2)
```

The main feature of `{getACS}` is support for returning multiple tables, geographies, and years.

```{r}
acs_data <- get_acs_geographies(
  geography = c("county", "state"),
  county = "Baltimore city",
  state = "MD",
  table = "B08134",
  year = 2022,
  quiet = TRUE
)
```

The package also includes utility functions for filtering data and selecting columns to support the creation of tables using the `{gt}` package:

```{r}
tbl_data <- filter_acs(acs_data, indent == 1, line_number <= 10)
tbl_data <- select_acs(tbl_data)

commute_tbl <- gt_acs(
  tbl_data,
  groupname_col = "NAME",
  column_title_label = "Commute time",
  table = "B08134"
)

as_raw_html(commute_tbl)
```

The `gt_acs_compare()` function also allows side-by-side comparison of geographies:

```{r}
commute_tbl_compare <- gt_acs_compare(
  data = tbl_data,
  id_cols = "column_title",
  column_title_label = "Commute time",
  table = "B08134"
)

as_raw_html(commute_tbl_compare)
```

`gt_acs_compare_vars()` is a variant on `gt_acs_compare()` where the default values support comparisons with values in columns and geographical areas in rows:

```{r}
commute_tbl_compare_vars <- acs_data |>
  filter_acs(indent == 1, line_number > 10) |>
  gt_acs_compare_vars(
    table = acs_data$table_id
  )

as_raw_html(commute_tbl_compare_vars)
```

The package also includes several functions to support creating plots with the `{ggplot2}` package including `geom_acs_col()` and `labs_acs_survey()`:

```{r}
plot_data <- acs_data |>
  filter_acs(indent == 1, line_number > 10) |>
  select_acs() |>
  fmt_acs_county(state = "Maryland")

plot_data |>
  ggplot() +
  geom_acs_col(
    fill = "NAME",
    position = "dodge",
    color = NA,
    alpha = 0.75,
    perc = TRUE,
    errorbar_params = list(position = "dodge", linewidth = 0.25)
  ) +
  scale_y_discrete("Means of transportation to work") +
  scale_fill_viridis_d("Geography") +
  labs_acs_survey(
    .data = acs_data
  ) +
  theme_minimal()
```

The `geom_acs_col()` function calls `geom_acs_errorbar()` (passing the `errorbar_params` argument as additional parameters) and `scale_x_acs()` or `scale_y_acs()` (depending on whether `orientation = "y"` or the default value of `NA`).

For more information on working with Census data in R [read the book *Analyzing US Census Data: Methods, Maps, and Models in R*](https://walker-data.com/census-r) (February 2023).

## Related projects

### Related R packages and analysis projects

-   [{easycensus}](https://github.com/CoryMcCartan/easycensus): Quickly Extract and Marginalize U.S. Census Tables
-   [{cwi}](https://ct-data-haven.github.io/cwi/): Functions to speed up and standardize Census ACS data analysis for multiple staff people at DataHaven, preview trends and patterns, and get data in more layperson-friendly
-   [{camiller}](https://camille-s.github.io/camiller/): A set of convenience functions, functions for working with ACS data via `{tidycensus}`
-   [{psrccensus}](https://psrc.github.io/psrccensus/): A set of tools developed for PSRC (Puget Sound Regional Council) staff to pull, process, and visualize Census Data for geographies in the Central Puget Sound Region.
-   [{CTPPr}](https://github.com/Westat-Transportation/CTPPr): A R package for loading and working with the US Census CTPP survey data.
-   [{lehdr}](https://github.com/jamgreen/lehdr): a package to grab LEHD data in support of city and regional planning economic and transportation analysis
-   [{mapreliability}](https://github.com/austensen/mapreliability): A R package for map classification reliability calculator
-   [Studying Neighborhoods With Uncertain Census Data](https://github.com/geoss/acs_demographic_clusters): Code to create and visualize demographic clusters for the US with data from the American Community Survey

### Related Python libraries

- [census-data-aggregator](https://github.com/datadesk/census-data-aggregator): A Python library from the L.A. Times data desk to help "combine U.S. census data responsibly"
- [census-table-metadata](https://github.com/censusreporter/census-table-metadata): Tools for generating metadata about tables and fields in a Census release based on sequence lookup and table shell files. (Note: the pre-computed data from this repository is used to label ACS data by `label_acs_metadata()`)
