---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# getACS

<!-- badges: start -->

[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental) [![License: MIT](https://img.shields.io/badge/License-MIT-yellow.svg)](https://opensource.org/licenses/MIT) [![Project Status: WIP -- Initial development is in progress, but there has not yet been a stable, usable release suitable for the public.](https://www.repostatus.org/badges/latest/wip.svg)](https://www.repostatus.org/#wip)

<!-- badges: end -->

The goal of getACS is to make it easier to work with [American Community Survey data](https://www.census.gov/programs-surveys/acs) from the [tidycensus package](https://walker-data.com/tidycensus/) by Kyle Walker and others.

This package includes:

-   Functions that extend `tidycensus::get_acs()` to support multiple tables, geographies, or years
-   Functions for creating formatted tables from ACS data using the [gt package](https://gt.rstudio.com/)

Note that I don't love the current name for this package and expect to rename it as soon as I think of a better one.

## Installation

You can install the development version of getACS from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pkg_install("elipousson/getACS")
```

## Usage

```{r setup}
library(getACS)
library(gt)
library(ggplot2)
```

The main feature of `{getACS}` is support for returning multiple tables, geographies, and years.

```{r}
acs_data <- get_acs_geographies(
  geography = c("county", "state"),
  county = "Baltimore city",
  state = "MD",
  table = "B08134",
  quiet = TRUE
)
```

The package also includes utility functions for filtering data and selecting columns to support the creation of tables using the `{gt}` package:

```{r}
tbl_data <- filter_acs(acs_data, indent == 1, line_number <= 10)
tbl_data <- select_acs_cols(tbl_data)

commute_tbl <- gt_acs(
  tbl_data,
  groupname_col = "NAME",
  column_title_label = "Commute time",
  table = "B08134"
)

as_raw_html(commute_tbl)
```

The package also includes several simple functions to support creating plots with the `{ggplot2}` package:

```{r}
plot_data <- filter_acs(acs_data, indent == 1, line_number > 10)
plot_data <- select_acs_cols(plot_data)

plot_data |>
  fmt_acs_county(state = "Maryland") |>
  ggplot(aes(x = perc_estimate, y = column_title, fill = NAME)) +
  geom_col(position = "dodge", alpha = 0.75) +
  scale_x_acs_percent() +
  scale_fill_viridis_d() +
  theme_minimal() +
  theme(legend.position = "top") +
  labs_acs_survey(
    y = "Means of transportation to work",
    fill = "Geography",
    table = acs_data$table_id
  )
```

For more information on working with Census data in R [read the book *Analyzing US Census Data: Methods, Maps, and Models in R*](https://walker-data.com/census-r) (February 2023).

## Related projects

### Related R packages

-   [{easycensus}](https://github.com/CoryMcCartan/easycensus): Quickly Extract and Marginalize U.S. Census Tables
-   [{cwi}](https://ct-data-haven.github.io/cwi/): Functions to speed up and standardize Census ACS data analysis for multiple staff people at DataHaven, preview trends and patterns, and get data in more layperson-friendly
-   [{camiller}](https://camille-s.github.io/camiller/): A set of convenience functions, functions for working with ACS data via `{tidycensus}`

### Related Python libraries

- [census-data-aggregator](https://github.com/datadesk/census-data-aggregator): A Python library from the L.A. Times data desk to help "combine U.S. census data responsibly"
- [census-table-metadata](https://github.com/censusreporter/census-table-metadata): Tools for generating metadata about tables and fields in a Census release based on sequence lookup and table shell files. (Note: the pre-computed data from this repository is used to label ACS data by `label_acs_metadata()`)
