---
title: "Introduction to getACS"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to getACS}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  message = FALSE,
  comment = "#>"
)
```

```{r setup}
library(getACS)
library(dplyr)
library(ggplot2)
theme_set(theme_minimal())
```

This package is designed to build on the `{tidycensus}` package to make it easier to create reproducible reports by allowing users to:

- Download multiple tables with `get_acs_tables()`
- Download tables for multiple geographies and tables with `get_acs_geographies()`
- Download tables for multiple years and tables with `get_acs_ts()`

All of these functions use the `label_acs_metadata()` to join the variables to detailed pre-computed [table and column metadata available on GitHub](https://github.com/censusreporter/census-table-metadata) from the open-source [Census Reporter project](https://censusreporter.org/). This metadata includes a parent column ID that supports the conversion of estimate and margin of error values into a percent share of the corresponding total. The function also uses the `race_iteration` reference to add a name of the racial or ethnic group as a column where appropriate.

## Using `get_acs_tables()`

```{r}
tenure_tables <- acs_tables_race_iteration("B25003")[c(2, 3, 10)]

tenure_data <- get_acs_tables(
  geography = "county",
  state = "MD",
  county = "Baltimore city",
  table = tenure_tables
)
```


```{r}
tenure_data |>
  filter_acs(indent > 0) |>
  select_acs_cols("race_iteration_group") |>
  ggplot(aes(x = column_title, y = perc_estimate, fill = race_iteration_group)) +
  geom_col(alpha = 0.8, position = "dodge") +
  scale_y_acs_percent("% of households") +
  scale_fill_viridis_d("Race/ethnic group") +
  labs_acs_survey(table = tenure_tables, x = "Tenure")
```

## Using `get_acs_geographies()`

```{r}
multigeo_acs_data <- get_acs_geographies(
  geography = c("county", "state"),
  state = "MD",
  county = "Baltimore city",
  table = "B01003"
)
```

```{r}
multigeo_acs_data |>
  select_acs_cols(column_title_col = NULL) |>
  gt_acs(
    rowname_col = "NAME",
    est_col_label = "Population"
  ) |>
  fmt_acs_county(state = "Maryland")
```

## Using `get_acs_ts()`

`get_acs_ts()` relies on a helper, `acs_survey_ts()`, that identifies the non-overlapping, comparable sample years for a specific ACS sample:

```{r}
years <- acs_survey_ts("acs5", 2021)

years
```

The `get_acs_ts()` calls `acs_survey_ts()` internally to return data for multiple years (warning you if a variable is unavailable for a specific year or geography):

```{r}
acs_ts_data <- get_acs_ts(
  geography = "county",
  state = "MD",
  survey = "acs5",
  year = 2021,
  table = "B01003"
)

glimpse(acs_ts_data)
```

Helper functions for `{ggplot2}` include `scale_x_acs_ts()` to set appropriate breaks:

```{r}
acs_ts_data |>
  filter_acs(
    GEOID %in% c("24510", "24005", "24003", "24027", "24025", "24035")
  ) |>
  ggplot(aes(x = year, y = estimate, fill = NAME)) +
  geom_col(alpha = 0.8) +
  facet_wrap(~NAME) +
  scale_y_acs_estimate("Population") +
  scale_x_acs_ts(survey = "acs5", year = 2021) +
  scale_fill_brewer(type = "qual", guide = "none")
```
